---
title: "Initial SJ markdown"
author: "WF"
date: "30 01 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The analysis below corresponds with the data uploaded yesterday (27.01.21) on kaggle.com. In this case we will not go beyond the basic EDA. I will mainly concentrate on the file containing information about competitions. 

## How these data look like?

Let's begin with the data and libraries import:

```{r import}
library(tidyverse)

competitions = read.csv2('all_comps.csv',sep=',',dec='.',stringsAsFactors = TRUE)
results = read.csv2('all_results.csv',sep=',',dec='.')
names = read.csv2('all_names.csv',sep=',',dec='.',stringsAsFactors = TRUE)

summary(competitions)
summary(results)
```
In fact in 'competitions' dataset we have mostly integer variables. What is interesting only from this code is that
* the NA values are apparent - mostly in these cases, when 
+ the wind/gate factor was not taken into account (roughly 25% of competitions and 20% of jumps, mainly in COC/FC competitions)
+ the style was not counted (all training rounds, roughly 45% of total number of jumps)
+ parsing of PDF file was not successful (roughly 6% of competitions, mainly before 2012/13 and for COC/FC types)
* most competitions was held in Germany and in Norway.

Let's see the most "fashioned" places and countries
```{r pressure, echo=FALSE}
require(ggplot2)
require(tidyverse)

counts <- as.data.frame(table(competitions$place,competitions$gender))
counts_raw <- counts[aggregate(Freq ~ Var1,counts, sum)$Freq>50,]
counts <- filter(counts,counts$Var1 %in% counts_raw$Var1)
counts %>%
  mutate(Var1 = fct_reorder(Var1,Freq,sum)) %>%
  ggplot(aes(x=Var1, y=Freq, fill=Var2)) +
    geom_bar(stat="identity", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    theme_bw()

countries <- as.data.frame(table(competitions$country,competitions$gender))
countries_raw <- countries[aggregate(Freq ~ Var1,countries, sum)$Freq>50,]
countries <- filter(countries,countries$Var1 %in% countries_raw$Var1)
countries %>%
  mutate(Var1 = fct_reorder(Var1,Freq,sum)) %>%
  ggplot(aes(x=Var1, y=Freq, fill=Var2)) +
    geom_bar(stat="identity", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    theme_bw()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
