---
title: "Initial SJ markdown"
author: "WF"
date: "17 02 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

The analysis below describes in quite detail competition placed on a large hill in Oberstdorf (hill size 137 meters). The analysis is connected with World Ski Championships, which are to be organized next week.

# Beginning

Lets begin with the data and libraries import:

```{r import}
library(tidyverse)
library(e1071)
library(shiny)
competitions = read.csv('all_comps_r.csv',sep=',',dec='.',stringsAsFactors = TRUE)
results = read.csv('all_results.csv',sep=',',dec='.')
names = read.csv('all_names.csv',sep=',',dec='.',stringsAsFactors = TRUE)
```

Actually, we do not have any interest in data not connected with LH in Oberstdorf, therefore we filter them:

```{r filter}
obe_ids = subset(competitions, place == 'Oberstdorf' & hill_size_x == 137)
actual_results <- subset(results, results$id %in% obe_ids$id)
summary(actual_results)
summary(obe_ids)
```
As we can see, Schattenbergschanze (the name of this hill) hosts mainly World Cup competitions and a few Continental Cup races. 

To see how the variables are correlated, we made a scatterplot and a correlation matrix:

```{r}
plot(actual_results[,c(1,2,9,11,13,15)])
round(cor(actual_results[,c(1,2,9,11,13,15)],use = 'complete.obs'),2)
```
This chunk suggests that

* the distance is positively correlated with style marks (+0.75) and wind (+0.3), but is negatively correlated with the gate (-0.28) and the speed (-0.06), what is surprising, but we will discuss it later,
* loc variable (place in given round) is negatively correlated with distance and style marks, what is quite obvious.
* wind is negatively correlated with the loc variable (-0.15), so this suggests (at least initially) that the jumpers with better wind have bigger chance to achieve a good result.

We can plot, how speed and distance varies

```{r}
ggplot(subset(actual_results[,c('dist','id')], dist>0)) +
  geom_density(aes(x = dist),fill="white", alpha = 0.6)+labs(title='Normalized plots of conditional densities')
ggplot(subset(actual_results[,c('speed','id')], speed>85)) +
  geom_density(aes(x = speed),fill="white", alpha = 0.6)+labs(title='Normalized plots of conditional densities')
ggplot(subset(actual_results[,c('speed','dist')], speed>85 & dist>0)) +
  geom_density_2d_filled(aes(x = speed, y= dist))+labs(title='Normalized bivariate distribution')
```

The last plot can be quite surprising -- is that real, that the higher/lower speed does not have an impact on distances, as this figure suggests? Of course that is not true, but we will comment that later. Now we will make a distinction between men and women.

```{r}
obe_men = subset(obe_ids, gender=='Men')$id
ggplot(subset(actual_results[,c('speed','dist','id')], speed>85 & dist>0 & id %in% obe_men)) +
  geom_density_2d_filled(aes(x = speed, y= dist))+labs(title='Normalized bivariate distribution for men') + xlim(87,100) + ylim(80,140)
obe_men = subset(obe_ids, gender=='Men')$id
ggplot(subset(actual_results[,c('speed','dist','id')], speed>85 & dist>0 & !(id %in% obe_men))) +
  geom_density_2d_filled(aes(x = speed, y= dist))+labs(title='Normalized bivariate distribution for women') + xlim(87,100) + ylim(80,140)
```

We can see that these conditional distributions are quite different and most "probable" distance varies from case to case. We can also explain why the correlation between the speed and distance is negative - simply bigger speed is connected with women rather than men and their distances are smaller. Later we will see that in fact the correlation is positive, if the gate is fixed.

Now we can examine the gate variable and see how the speed changes along with the higher/lower start position.

```{r gate}
ui <- basicPage(
  selectInput("gate", label = 'Select gate:',
  choices = sort(unique(actual_results$gate)), selected = 15
  ),
  plotOutput("den")
)

server <- function(input, output) {
  output$den <- renderPlot({
  gate_results  = subset(merge(actual_results[,c('speed','id','gate')],competitions[,c('id','gender')],by='id'), gate == input$gate & speed > 85)
  N_1 = nrow(subset(gate_results, gender=='Women'))
  N = paste('Women jumps count: ', N_1, ', Men jumps count: ', nrow(gate_results) - N_1)
  ggplot(merge(gate_results[,c('speed','id')],competitions[,c('id','gender')],by='id')) +
  geom_histogram(aes(x = speed, color = gender),fill="white", alpha = 0.6)+labs(title=N) + xlim(87.5,97.5)
})
}

shinyApp(ui, server)
```

The results are weird, because the distribution is not uni-modal. The explanation is simple -- the labelling of start position changed in the mid-time. To see, when (probably) this change occurred.

```{r gate_shift}
ui <- basicPage(
  selectInput("gate", label = 'Select gate:',
  choices = sort(unique(actual_results$gate)), selected = 15
  ),
  selectInput("year", label = 'Select a limit year:',
  choices = seq(2010,2021,by = 1), selected = 2018
  ),
  plotOutput("den")
)

server <- function(input, output) {
  output$den <- renderPlot({
  gate_results  = subset(merge(actual_results[,c('speed','id','gate')],competitions[,c('id','season')],by='id'), gate == input$gate & speed > 85)
  gate_results['before'] = cut(gate_results$season, breaks = c(2009,input$year,2022), labels=c('to', 'after')) 
  N = paste('Before and after: ', input$year)
  ggplot(gate_results) +
  geom_histogram(aes(x = speed, color = before),fill="white", alpha = 0.6) + labs(title=N) + xlim(87.5,97.5)
})
}

shinyApp(ui, server)
```

We conclude that every competition after (i.e. not including) 2018/2019 WC season has "new" way of labelling. Moreover, we can estimate the difference in two ways:

* Now the same number of start position means 1-1.5 lower inrun speed (in kmh) or
* The same speed now is achieved by taking higher start position by 3-4 levels. 

To make the further part clear, we will not include the gate variable, but operate only on the speed variable. Although it increases variance (athletes can achieve different velocity from the same inrun length), it should not affect the results significantly.

Now we can check how the distances depend on the speed and the wind variable:

```{r}
ui <- basicPage(
  selectInput("gender", label = 'Select gender:',
  choices = unique(competitions$gender), selected = "Men"
  ),
  selectInput("year", label = 'Select a limit year:',
  choices = seq(2010,2021,by = 1), selected = 2018
  ),
  plotOutput("den")
)

server <- function(input, output) {
  output$den <- renderPlot({
  gate_results  = subset(merge(actual_results[,c('speed','id', 'gate', 'wind', 'dist')],
                               competitions[,c('id', 'season', 'gender')],by = 'id'), 
                         season == input$year & gender == input$gender & speed > 85 & dist > 0)
  N = paste('Results from : ', input$year, ' Gender:', input$gender)
  ggplot(gate_results) +
  geom_raster(aes(x = speed, y = dist, fill = wind), alpha = 0.6) + labs(title=N) + xlim(87.5,97.5) + ylim(90,145)
})
}
shinyApp(ui, server)
```

