---
title: "Ratings insights"
author: "Wiktor Florek"
date: "24 03 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

The analysis below corresponds with the data uploaded recently (24.03.21) on kaggle.com. We have a new dataset called all_ratings.csv. We will take care of the content included inside this dataframe.

# How ratings data look like?

Lets begin with the data and libraries import:

```{r import}
library(tidyverse)
library(e1071)
ratings = read.csv('all_ratings.csv',sep=',',dec='.',stringsAsFactors = TRUE)
names = read.csv('all_names.csv',sep=',',dec='.',stringsAsFactors = TRUE)
competitions = read.csv('all_comps_r.csv',sep=',',dec='.',stringsAsFactors = TRUE)
results = read.csv('all_results.csv',sep=',',dec='.',stringsAsFactors = TRUE)

ratings['performance'] = 2*as.numeric(cut(
    ratings$delty,
    breaks = c(-1000,0,1000),
    labels = c(-1,+1)
  ))-3

summary(ratings)
```

The main "star" of this dataset is "delty" variable. This vector informs about *relative* performance of every athlete in a given round. For example, if we have two athletes with ratings 1600 and 1300 respectively and the first is worse that the latter, then delty variable is highly negative in the first case and highly positive in the second case. If the 1600-rating player wins with the 1300-rating player, then delty variable is mildly positive in the first case and mildly positive in the second case. The baseline ratings are stored in *cumm_rating* variable for each competition. 

# First insights

We can now see, how this rating reflexes the performance of every athlete. For that we will need some instance of a rating graph:

```{r rating_graph_1}
graph_names = c('kraft stefan','kobayashi ryoyu', 'kasai noriaki', 'granerud halvor egner', 'loitzl wolfgang')

for (name in graph_names){
  jumper_id = names[names['name'] == name,2][1]

  filtered_ratings = filter(ratings, codex == jumper_id)[, c('cumm_rating', 'id', 'number')]
  tmp_title <- paste('Performance graph of: ', name)
  graph <- ggplot(filtered_ratings, aes(number, cumm_rating)) + geom_smooth() + geom_line() + labs(title=tmp_title) + xlim(0,max(ratings['number'])) + ylim(600,2000)
  print(graph)
}
```

Therefore we can see that this rating system quite well describes the level of particular athletes. We can also separate the delty variable and group it by the type of the hill. Then we are able to see which athlete is performing well on each type of hill.

```{r rating_graph_by_hill}
graph_names = c('kraft stefan',
                'kubacki dawid',
                'kasai noriaki',
                'lundby maren')

for (name in graph_names) {
  jumper_id = names[names['name'] == name, 2][1]
  
  filtered_ratings = filter(ratings, codex == jumper_id)[, c('delty', 'id', 'number')]
  filtered_ratings = merge(filtered_ratings,
                           competitions,
                           by.x = 'id',
                           by.y = 'id')[, c('delty', 'id', 'meter.value', 'number')]
  filtered_ratings['size'] = cut(
    filtered_ratings$meter.value,
    breaks = c(0, 1.3, 1.9, 2.3, 2.5),
    labels = c('flying hill', 'large', 'normal', 'medium')
  )
  filtered_ratings$csum <-
    ave(filtered_ratings$delty, filtered_ratings$size, FUN = cumsum)
  filtered_ratings = merge(
    filtered_ratings,
    as.data.frame(table(filtered_ratings$size)),
    how = 'left',
    by.x = 'size',
    by.y = 'Var1'
  )
  tmp_title <- paste('Performance boxplot by hill of: ', name)
  print(ggplot(filtered_ratings, aes(size, delty, fill = size)) + geom_boxplot() + ylim(-20, 20) + labs(title=tmp_title))
  tmp_title_2 <- paste('Performance graph by hill type of: ', name)
  print(ggplot(filtered_ratings, aes(number, csum, color = size)) + geom_smooth() + geom_line() + labs(title=tmp_title_2)+ xlim(0,max(ratings['number'])) + ylim(-500,1000))
}
```

These graphs suggests that indeed some athletes perform much better on normal hills, some on flying hills. Now we will create a barplot with the top athletes to see which of them correspond to which category.

```{r top_athletes}
selected_ids = unique(filter(competitions,season == 2021, gender == 'Men')[['id']])
selected_codex = unique(filter(ratings, id %in% selected_ids & cumm_rating > 1550)[['codex']])
selected_ratings = filter(ratings, codex %in% selected_codex)
selected_ratings = merge(selected_ratings,
                           competitions,
                           by.x = 'id',
                           by.y = 'id')[, c('id', 'performance', 'meter.value', 'codex.x')]
selected_ratings['size'] = cut(
    selected_ratings$meter.value,
    breaks = c(0, 1.3, 1.9, 2.3, 2.5),
    labels = c('flying hill', 'large', 'normal', 'medium')
  )
size_ratings = selected_ratings %>% group_by(size,codex.x) %>% summarize(performance = mean(performance, na.rm = TRUE)) %>% na.omit()
size_ratings = merge(size_ratings,names[!duplicated(names[,c('codex')]),], how='left', by.x='codex.x', by.y='codex')
tmp_title <- 'Performance on all types of hills'
size_ratings %>% arrange(name) %>% ggplot(aes(name,performance, fill = size)) + geom_bar(stat='identity') + labs(title=tmp_title)+coord_flip()
```

We observe that all top jumpers are quite good on large hills. In the other cases it can vary quite substantially -- some of them prefer normal hills, some of them are good on flying hills and the most surprising case is *Junshiro Kobayashi*, who jumps well only on large hills and loses on both normal and flying hills. Now we will provide a graph with top athletes on each type of hill.

```{r top_by_hill_type}
selected_ids = unique(filter(competitions,season == 2021, gender == 'Men')[['id']])
selected_codex = unique(filter(ratings, id %in% selected_ids)[['codex']])
selected_ratings = filter(ratings, codex %in% selected_codex)
selected_ratings = merge(selected_ratings,
                           competitions,
                           by.x = 'id',
                           by.y = 'id')[, c('id', 'performance', 'meter.value', 'codex.x')]
selected_ratings['size'] = cut(
    selected_ratings$meter.value,
    breaks = c(0, 1.3, 1.9, 2.3, 2.5),
    labels = c('flying hill', 'large', 'normal', 'medium')
  )
size_ratings = selected_ratings %>% group_by(size,codex.x) %>% summarize(performance = mean(performance, na.rm = TRUE), n = n()) %>% na.omit()
size_ratings = merge(size_ratings,names[!duplicated(names[,c('codex')]),], how='left', by.x='codex.x', by.y='codex')
for (hill_type in levels(size_ratings$size)){
  u=table(levels(size_ratings$size))
  u[c('large','flying hill','normal','medium')]=c(90,30,60,10)
  size_ratings_type = filter(size_ratings, size == hill_type & performance > 0 & n>=u[hill_type])
  size_ratings_type = size_ratings_type %>% arrange(desc(performance)) %>% slice(1:15)
  tmp_title <- paste('Performance on', hill_type)
  print(size_ratings_type %>% mutate(name=fct_reorder(name,performance,sum)) %>% ggplot(aes(name,performance, fill = size)) + geom_bar(stat='identity') + labs(title=tmp_title)+coord_flip())
}
```

We can also check the performance by other factors like wind factor. The graph below says "Which athletes are better when the wind is disturbing or helping"

```{r top_by_wind}
selected_codex = unique(filter(ratings, id %in% selected_ids & cumm_rating > 1500)[['codex']])
selected_ratings = filter(ratings, codex %in% selected_codex)
selected_ratings = left_join(selected_ratings, results, by = c('codex','round','id'))[,c('codex','performance','delty','wind')] %>% na.omit()
selected_ratings['wind_type'] = cut(selected_ratings$wind, breaks = c(-10,0,10), labels=c('tail','head'))
wind_ratings = selected_ratings %>% group_by(wind_type,codex) %>% summarize(performance = mean(performance, na.rm = TRUE), n = n()) %>% na.omit()
wind_ratings = merge(wind_ratings,names[!duplicated(names[,c('codex')]),], how='left', by.x='codex', by.y='codex')
tmp_title <- 'Performance while tail or head wind'
wind_ratings %>% mutate(name=fct_reorder(name,performance,sum)) %>% ggplot(aes(name,performance, fill=wind_type)) + geom_bar(stat='identity')+ coord_flip() + labs(title=tmp_title)
```

We can observe that most of the top athletes deals quite well regardless of the type of wind. However, some are more sensitive to tail wind, in particular Norwegians. The other extreme is when a jumper is better, whenever tail wind is present. We can see that *Dawid Kubacki* or *Andrzej Stêka³a* can be considered as this type of jumper.

Now we will analyze the *peaks* of ratings of particular jumpers.

```{r peaks}
peak_ratings = ratings %>% arrange(desc(cumm_rating)) %>% distinct(codex, .keep_all = TRUE)  %>% slice(1:30)
peak_ratings = merge(peak_ratings,
                           competitions,
                           by.x = 'id',
                           by.y = 'id')[, c('id', 'cumm_rating', 'date', 'codex.x')]
peak_ratings = merge(peak_ratings,names[!duplicated(names[,c('codex')]),], how='left', by.x='codex.x', by.y='codex')
peak_ratings['year'] = sapply(peak_ratings['date'],substring,1,4)
peak_ratings['year'] = sapply(peak_ratings['year'],as.numeric)
tmp_title <- paste('Peak ratings')
peak_ratings %>% mutate(name=fct_reorder(name,cumm_rating,sum)) %>% ggplot(aes(name,cumm_rating, fill=year)) + geom_bar(stat='identity') + labs(title=tmp_title)+coord_flip(ylim = c(1600, 2000))+scale_fill_gradient(low='red',high='green')
```

From this chart we can see that ratings over 1700 are impressive and not many jumpers (either men or women) can achieve that level of mastery. The record so far is approximately 1870 points and is hold by *Stefan Kraft* since March 2020.

# Conclusions

We can see that the ratings dataset can be used in various ways. We can obtain which athletes are good whenever some conditions are given. In fact this simple system can produce a lot of interesting analysis and this one is only an example. You're invited to try some of these on your own. 

# Experiments

```{r}
year=2014
selected_ids = unique(filter(competitions,season == year, gender == 'Men')[['id']])
selected_codex = unique(filter(ratings, id %in% selected_ids)[['codex']])
selected_ratings = filter(ratings, codex %in% selected_codex)
selected_ratings = left_join(selected_ratings, competitions, by = c('id'))[,c('codex.x','season','cumm_rating')]
selected_ratings = filter(selected_ratings, season < year + 1)
selected_ratings = selected_ratings[!duplicated(selected_ratings$codex.x, fromLast = TRUE),]
selected_ratings  %>% ggplot(aes(cumm_rating, y=..density..)) + geom_density() + xlim(500,2000) + ylim(0,0.007)
```

